Films: Age & Gender

### 2017-08-22 01_actor_careers/05_report.Rmd
fixes and theme added

---
title: "Test RMD"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Intro

Stuff and text.

## Data

### Films

```{r}
library(readr)
library(dplyr)
library(tidyr)

@@

### Actors / Directors

@@

# library(gender)
# library(stringr)
# library(lubridate)
# 
# film_folk = read_csv("https://raw.githubusercontent.com/TMBish/lab_posts/master/01_actor_careers/00_actors_directors.csv")
# 
# film_folk[film_folk$dob < as.Date("1880-01-01"), "dob"] = as.Date("1880-01-01")
# 
# # Grep out first name and dob year for the gender function
# film_folk = 
#   film_folk %>%
#   mutate(
#     first_name = str_extract(name, "(?i)[a-z]+(?=\\s)"),
#     dob_year = year(dob)
#   )
#   
# # Distinct name, dob combos and predict gender
# genders = 
#   film_folk %>% 
#   distinct(first_name, dob_year) %>%
#   gender_df(name_col = "first_name", year_col = "dob_year") %>%
#   select("first_name" = name, gender, "dob_year" = year_min)

@@

We should sense check the gender mapping. Let's take a random 10 people to check:

```{r}
genders %>%
  sample_n(10) %>%
  knitr::kable(caption = "Sense Checking the Gender Mapping")

@@

Looks pretty, pretty, pretty good.

@@

## Gender Gap

Some text.
Some text.
Some text.
Some text.Some text.
Some text.
Some text.
Some text.
Some text.
Some text.
Some text.

```{r}
library(highcharter)
tmbish = 
  hc_theme(
    chart = list(
      backgroundColor = list("white"),
      style = list(
        fontFamily = "Droid Sans"
      )
    ),
    colors = list("#414141", "#FF6A5C", "#CCDFC8", "#056571"),
    title = list(
      style = list(fontWeight = "bold", color = "#414141"),
      align = "left"
    ),
    xAxis = list(
      gridLineColor = "#2196F3",
      lineColor = "#2196F3",
      gridLineWidth = 0.5,
      lineWidth = 0,
      title = list(style = list(fontWeight = "bold"))
    ),
    yAxis = list(
      gridLineColor = "#2196F3",
      lineColor = "#2196F3",
      gridLineWidth = 0.5,
      lineWidth = 0,
      title = list(style = list(weight = "bold"))
    ),
    subtitle = list(
      style = list(fontStyle = "italic", color ="#414141"), 
      align = "left"
    ),
    tooltip = list(
      borderWidth = 0,
      shape = "square",
      valueDecimals = 2
    )
  )

volume_data =
  films %>% 
  select(title, year, tomatometer, director, actor_1:actor_6) %>%
  gather("role", "name", director, actor_1:actor_6) %>%
  mutate(role = ifelse(role == "director", "Director", "Actor")) %>%
  inner_join(film_folk) %>%
  mutate(
    age_at_production = year - dob_year
  ) %>%
  filter(between(age_at_production, 10,100)) %>%
  select(name, gender, dob_year, age_at_production, year, role, tomatometer)

  temp = 
    volume_data %>%
    filter(
      !is.na(gender)
    ) %>%
    mutate(
      category = case_when(
        role == "Director" ~ "director",
        gender == "male" ~ "male_actor",
        TRUE ~ "female_actor"
      )
    ) %>% 
    group_by(category, age_at_production) %>%
    summarise(films = n()) %>%
    group_by(category) %>%
    mutate(percentage = films / sum(films))

@@

<div id = "age_series_chart">
```{r}
chart_data = 
temp %>%
select(-films) %>%
spread(category, percentage) %>%
mutate(
  director = ifelse(!is.na(director), director, 0),
  female_actor = ifelse(!is.na(female_actor), female_actor, 0),
  male_actor = ifelse(!is.na(male_actor), male_actor, 0),
  director_cdf = cumsum(director),
  female_actor_cdf = cumsum(female_actor),
  male_actor_cdf = cumsum(male_actor)
)

convert_js_vec = function(vec) {

  js_string = vec %>% round(4) %>% paste(collapse = ",")

  return(JS( paste0("[", js_string, "]") ))

}

@@

chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(
    animation = list(duration = 2000)
  ) %>%
  hc_xAxis(
    categories = chart_data$age_at_production,
    title = list(text = "Age at Production")
  ) %>%
  hc_yAxis(
    title = list(text = "Probablity Density"),
    labels = list(formatter = JS("function(){return(Math.round(this.value * 1000) / 10 + '%')}"))
  ) %>%
  hc_title(text = "Female actors") %>%
  hc_subtitle(text = "Comparing male actors, (male) directors and female actors") %>%
  hc_add_series(
    name = "Director", 
    type = "spline",
    data = chart_data$director,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Male Actor",
    type = "spline",
    data = chart_data$male_actor,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Female Actor",
    type = "area",
    data = chart_data$female_actor,
    zIndex = -10,
    marker = list(enabled = FALSE),
    fillOpacity = 0.5
  )

chart
```
</div>

<button id = "update_age_series"> Show Cumulative Distribution </button>

@@

To investigate our hypothesis we must make a couple of assumptions. Since we don't have an individual actor / director performance (like we would have for players in a sporting contest for example) we'll have to assume an actor or director can have a tangible affect on the quality of the movie they appear in; there are obviously many other factors but this seems like a relatively safe assumption.

Further - for actors - we'll have to limit the dataset to actors that played an important role in the film. They  need to be prominent enough so as to affect the overall quality of the film. So while Michael Cera's cameo in *This is the End* was quality it can't be said that he influenced the rotten tomatoes rating of the film on his own.

@@

After looking through a few rotten tomatoes pages it seems limiting to the top 3 actors in a given film is a good rule of thumb for at least partially influncial performances.

### Average Performance by Age

## Housekeeping

@@

          $("#update_age_series").html("Show Density");
        };

>>>>

---
title: "Films: Age & Gender"
output: 
html_document:
code_folding: hide
---

# Questions?

Let's say we had a couple of questions about actors and directors:

> Do actors and directors improve with age?

> Are female actors generally cast  than their male counterparts?

To answer these questions we're going to need some data.

# Housekeeping

Let's set ourselves up with some packages, themes and utility functions.

```{r}
# Packages
library(readr)
library(dplyr)
library(tidyr)
library(gender)
library(stringr)
library(lubridate)
library(highcharter)

# Rmarkdown Options
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

# Convert an R vector to string JS vector
convert_js_vec = function(vec) {
  js_string = vec %>% round(4) %>% paste(collapse = ",")
  return(JS( paste0("[", js_string, "]") ))
}

# Highcharts theme
source()
tmbish = high_tmbish()
```

# Data

I started with imdb user **hagennelson's** list of [top 1000 actors](http://www.imdb.com/list/ls058011111/) and other imdb user **earthanime's** list of [top 250 directors](http://www.imdb.com/list/ls062419846/). Both lists span across generations, from the ancient (Ethel Barrymore and Alfred Hitchcock) to the modern (Chloë Grace Moretz and Christopher Nolan).

I then developed code (which I'll cover in a different post) to scrape data from the rotten tomatoes pages of the above individuals' films. The final dataset includes **12,258** films and **1,233** indviduals.

Altough we'll only use a few of these variables I brought in a few extras for some future investigations:

Variable | Description
-------- | -----------
url | URL extension on rotten tomatoes website
year | Film production year
title | Film title
reviews | Count of formal critic reviews of film
tomatometer | Percentage of favourable reviews
av_rating | Average rating of film out of 10
audience | The non-critic plebian equivolent of the tomatometer
box_office | The box-office revenue
director | The film's director
actor_1:actor_6 | The 6 top actors according to rotten tomatoes internal ranking (seems like screen time)

## Film Data

We'll load and take a peek at the film dataset:

@@

## Individual Data

Now we'll load the individual data.

Unfortunately, because the Rotten Tomatoes devs are a bunch of politically correct SJWs, gender isn't provided on their film bio pages so we'll have to be creative. The `gender` package in R allows you to infer a person's gender from their first name and DOB. The result is a probabilistic best guess of that individuals most likely gender given their first name and DOB.

@@

# library(gender)
# library(stringr)
# library(lubridate)
# 
# film_folk = read_csv("https://raw.githubusercontent.com/TMBish/lab_posts/master/01_actor_careers/00_actors_directors.csv")
# 
# # Clean some data quality issues on the DOB
# film_folk[film_folk$dob < as.Date("1880-01-01"), "dob"] = as.Date("1880-01-01")
# 
# # Grep out first name and dob year for the gender function
# film_folk =
#   film_folk %>%
#   mutate(
#     first_name = str_extract(name, "(?i)[a-z]+(?=\\s)"),
#     dob_year = year(dob)
#   )
# 
# # Distinct name, dob combos and predict gender
# genders =
#   film_folk %>%
#   distinct(first_name, dob_year) %>%
#   gender_df(name_col = "first_name", year_col = "dob_year") %>%
#   select("first_name" = name, gender, "dob_year" = year_min)

@@

We should sense check the gender mapping. Let's take a random 10 people to check is working as desired:

```{r}
set.seed(123)

genders %>%
  sample_n(10) %>%
  knitr::kable(caption = "Sense Checking the Gender Mapping")

@@

Things get dicey with a name like Sandy; however, the package obviously knows that most Sandys born in 1937 were female so we'll have to trust it here. Otherwise the list looks pretty, pretty, pretty good.

@@

# Q1. Gender Gap

To answer this question we'll first need to calculate a person's age at the time of their appearance in a film. We'll then

# Chain a bunch of data wrangling operations
volume_data =
  films %>% 
  select(title, year, tomatometer, director, actor_1:actor_6) %>%
  gather("role", "name", director, actor_1:actor_6) %>%
  mutate(role = ifelse(role == "director", "Director", "Actor")) %>%
  inner_join(film_folk) %>%
  mutate(
    age_at_production = year - dob_year
  ) %>%
  filter(
    between(age_at_production, 10,100),
    !is.na(gender)
  ) %>%
  select(name, gender, dob_year, age_at_production, year, role, tomatometer) %>%
  mutate(
    category = case_when(
      role == "Director" ~ "director",
      gender == "male" ~ "male_actor",
      TRUE ~ "female_actor"
    )
  ) %>% 
  group_by(category, age_at_production) %>%
  summarise(
    films = n()
  ) %>%
  group_by(category) %>%
  mutate(
    percentage = films / sum(films)
  )

@@

<div id = "age_series_chart">
```{r}

# Shape data for presentation
chart_data = 
  volume_data %>%
  select(-films) %>%
  spread(category, percentage) %>%
  mutate(
    director = ifelse(!is.na(director), director, 0),
    female_actor = ifelse(!is.na(female_actor), female_actor, 0),
    male_actor = ifelse(!is.na(male_actor), male_actor, 0),
    director_cdf = cumsum(director),
    female_actor_cdf = cumsum(female_actor),
    male_actor_cdf = cumsum(male_actor)
  )

@@

# Create chart
age_dist_chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(
    animation = list(duration = 2000)
  ) %>%
  hc_xAxis(
    categories = chart_data$age_at_production,
    title = list(text = "Age at Production")
  ) %>%
  hc_yAxis(
    title = list(text = "Probablity Density"),
    labels = list(formatter = JS("function(){return(Math.round(this.value * 1000) / 10 + '%')}"))
  ) %>%
  hc_title(text = "Female actors") %>%
  hc_subtitle(text = "Comparing male actors, (male) directors and female actors") %>%
  hc_add_series(
    name = "Director", 
    type = "spline",
    data = chart_data$director,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Male Actor",
    type = "spline",
    data = chart_data$male_actor,
    marker = list(enabled = FALSE)
  ) %>%
  hc_add_series(
    name = "Female Actor",
    type = "area",
    data = chart_data$female_actor,
    zIndex = -10,
    marker = list(enabled = FALSE),
    fillOpacity = 0.5
  )

age_dist_chart
```
</div>

<button class="button secondary outline" id = "update_age_series"> Show Cumulative Distribution </button>

# Q2. Acting and Directing as a skill

> Hypothesis 2: acting and directing skills aren't tied to a person's physical attributes so shouldn't experience deterioration in age. In fact you would expect these artists to improve their skills with age.

@@

To investigate our hypothesis we must make a couple of assumptions. Since we don't have an individual actor / director performance (like we have in sports) we'll have to assume an actor or director can have a tangible affect on the quality of the movie they appear in; there are obviously many other factors but this seems like a relatively safe assumption.

Further - for actors - we'll have to limit the dataset to actors that played an important role in the film. To implement an `actor performance == film quality` approach the actors need to be prominent enough so as to be able to affect the overall quality of the film. So while Michael Cera's cameo in **This is the End** was high quality it can't be said that he influenced the rotten tomatoes rating of the film on his own.

@@

After looking through a few rotten tomatoes pages it seems limiting to the top 3 actors in a given film is a reasonable cut-off for influncial performances.

> INSERT EXAMPLES FROM A FEW ROTTEN TOMATOES PAGES

## Methodology

For every person in the dataset we follow the following recipe:

1. Produce a dataset for each actor with films that:
+ We have a tomatometer score for
+ And the actor is in the top 3 of rotten tomatoes credits
2. Calculate the `mean` tomatometer for all of this person's films;
3. Calculate the age the person was at the release of each film;
4. For every film calculate what `percentage %` this tomato rating is above or below this person's `mean` (called a tomato index)

Producing a dataset like this:

```{r echo=FALSE}
films %>% 
  select(title, year, tomatometer, director, actor_1:actor_3) %>%
  gather("role", "name", director, actor_1:actor_3) %>%
  inner_join(film_folk) %>%
  filter(
    name == "Denzel Washington"
  ) %>%
  mutate(
    age_at_production = year - dob_year
  ) %>%
  filter(between(age_at_production, 15,85)) %>%
  select(title, name, year, gender, age_at_production, tomatometer) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = round(mean(tomatometer))
  ) %>% 
  ungroup() %>%
  mutate(
    tomato_index = tomatometer / average_tomato
  ) %>%
  arrange(age_at_production) %>%
  select(title, name, year, age_at_production, tomatometer, average_tomato, tomato_index) %>%
  mutate(index = scales::percent(tomato_index)) %>%
  sample_n(10) %>%
  knitr::kable()

## Results

```{r}
# 
rating_set =
  films %>% 
  select(title, year, tomatometer, director, actor_1:actor_3) %>%
  gather("role", "name", director, actor_1:actor_3) %>%
  inner_join(film_folk) %>%
  filter(!is.na(gender)) %>%
  mutate(
    age_at_production = year - dob_year,
    category = case_when(
      role == "director" ~ "director",
      gender == "male" ~ "male_actor",
      TRUE ~ "female_actor"
    )
  ) %>%
  filter(between(age_at_production, 15,85)) %>%
  select(name, gender, age_at_production, category, tomatometer) %>%
  group_by(name) %>% 
  mutate(
    average_tomato  = median(tomatometer),
    films = n()
  ) %>% ungroup() %>%
  filter(films >= 5)

rating_volumes = 
  rating_set %>%
  group_by(age_at_production) %>%
  summarise(films = -1*n())

rating_scores = 
  rating_set %>%
  mutate(
    tomato_index = tomatometer / average_tomato) %>%
  group_by(category, age_at_production) %>%
  summarise(rating = median(tomato_index)) %>%
  spread(category, rating)

# LOESS Trend Line
loess_fit = loess(value ~ age_at_production , 
                  data = rating_scores %>% gather("role", "value", -age_at_production)
                  # %>% group_by(age_at_production) %>% summarise(value = mean(value, na.rm = TRUE))
)

predictions = loess_fit %>% predict(seq(15,85))

volume_chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(height = "300") %>%
  hc_xAxis(categories = rating_volumes$age_at_production, opposite = TRUE) %>%
  hc_yAxis(
    title = list(text = "Total Films in Dataset"),
    labels = list(formatter = JS("function(){return(-1 * this.value)}"))    
    ) %>%
  hc_title(text = "") %>%
  hc_add_series(
    name = "Volumes",
    type = "area",
    data = rating_volumes$films,
    marker = list(enabled = FALSE)
  ) %>% 
  hc_legend(enabled = FALSE)

rating_chart = 
  highchart() %>%
  hc_add_theme(tmbish) %>%
  hc_chart(
    animation = list(duration = 2000)
  ) %>%
  hc_xAxis(
    categories = rating_scores$age_at_production,
    title = list(text = "Age at Production")
  ) %>%
  hc_yAxis(
    title = list(text = "Probablity Density"),
    min = 0.6
  ) %>%
  hc_title(text = "Female actors") %>%
  hc_subtitle(text = "Comparing male actors, (male) directors and female actors") %>%
  hc_add_series(
    name = "Director",
    type = "scatter",
    data = rating_scores$director,
    marker = list(radius = 3, symbol = "circle")
  ) %>%
  hc_add_series(
    name = "Male Actor",
    type = "scatter",
    data = rating_scores$male_actor,
    marker = list(radius = 3, symbol = "circle")
  ) %>%
  hc_add_series(
    name = "Female Actor",
    type = "scatter",
    data = rating_scores$female_actor,
    marker = list(radius = 3, symbol = "circle")
  ) %>%
  hc_add_series(
    name = "Trend",
    type = "spline",
    data = predictions,
    color = "#000000",
    marker = list(enabled = FALSE)
  )

```{r echo = FALSE}
chart
```

```{r echo = FALSE}
volume_chart
```

@@

$("#update_age_series").html("Show Density");

};
